---
title: "class18"
format: pdf
---

**Q1. With the help of the R “addin” package datapasta assign the CDC pertussis case number data to a data frame called cdc and use ggplot to make a plot of cases numbers over time.**

```{r}
cdc <- data.frame(
                                           Year = c(1922L,1923L,1924L,1925L,
                                                    1926L,1927L,1928L,1929L,
                                                    1930L,1931L,1932L,1933L,
                                                    1934L,1935L,1936L,1937L,
                                                    1938L,1939L,1940L,1941L,
                                                    1942L,1943L,1944L,1945L,
                                                    1946L,1947L,1948L,1949L,
                                                    1950L,1951L,1952L,1953L,
                                                    1954L,1955L,1956L,1957L,
                                                    1958L,1959L,1960L,1961L,
                                                    1962L,1963L,1964L,1965L,
                                                    1966L,1967L,1968L,1969L,
                                                    1970L,1971L,1972L,1973L,
                                                    1974L,1975L,1976L,1977L,
                                                    1978L,1979L,1980L,1981L,
                                                    1982L,1983L,1984L,1985L,
                                                    1986L,1987L,1988L,1989L,
                                                    1990L,1991L,1992L,1993L,
                                                    1994L,1995L,1996L,1997L,
                                                    1998L,1999L,2000L,2001L,
                                                    2002L,2003L,2004L,2005L,
                                                    2006L,2007L,2008L,2009L,
                                                    2010L,2011L,2012L,2013L,
                                                    2014L,2015L,2016L,2017L,
                                                    2018L,2019L,2020L,2021L),
                   No..Reported.Pertussis.Cases = c(107473,164191,165418,
                                                    152003,202210,181411,161799,
                                                    197371,166914,172559,
                                                    215343,179135,265269,180518,
                                                    147237,214652,227319,
                                                    103188,183866,222202,191383,
                                                    191890,109873,133792,
                                                    109860,156517,74715,69479,
                                                    120718,68687,45030,37129,
                                                    60886,62786,31732,28295,
                                                    32148,40005,14809,11468,
                                                    17749,17135,13005,6799,
                                                    7717,9718,4810,3285,4249,
                                                    3036,3287,1759,2402,1738,
                                                    1010,2177,2063,1623,
                                                    1730,1248,1895,2463,2276,
                                                    3589,4195,2823,3450,4157,
                                                    4570,2719,4083,6586,
                                                    4617,5137,7796,6564,7405,
                                                    7298,7867,7580,9771,11647,
                                                    25827,25616,15632,10454,
                                                    13278,16858,27550,18719,
                                                    48277,28639,32971,20762,
                                                    17972,18975,15609,18617,
                                                    6124,2116)
                 )
```
```{r}
library(ggplot2)
p <- ggplot(cdc) + aes(x=Year, y = No..Reported.Pertussis.Cases) + geom_point() + geom_line() + labs(x="Year", y="No. Reported Pertussis Cases", title="Pertussis Cases by Year (1922-2019)")
p
```

**Q2. Using the ggplot geom_vline() function add lines to your previous plot for the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine (see example in the hint below). What do you notice?**

```{r}
p <- p + geom_vline(xintercept = c(1946, 1996))
p
```

**Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?**

After introduction of the wP vaccine, there was a drastic drop in pertussis cases. The introduction of the aP vaccine produced a small erratic increase in cases, possibly due to more sensitive technology in detecting cases, growing public concern over vaccines, and vaccine immunity.

```{r}
library(jsonlite)
subject <- read_json("https://www.cmi-pb.org/api/subject", simplifyVector = TRUE)
head(subject, 3)
```
**Q4. How many aP and wP infancy vaccinated subjects are in the dataset?**

```{r}
table(subject$infancy_vac)
```
60 aP and 58 wP subjects

**Q5. How many Male and Female subjects/patients are in the dataset?**

```{r}
table(subject$biological_sex)
```
79 female, 39 male subjects

**Q6. What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc…)?**

```{r}
table(subject$biological_sex, subject$race)
```
1 American Indian/Alaska Native male, 0 American Indian/Alaska Native females, 21 Asian females, 11 Asian males, 2 Black / African American females, 0 Black / African American males, 9 females more than one race, 2 males more than one race, 1 Native Hawaiian or Other Pacific Islander female, 1 Native Hawaiian or Other Pacific Islander male, 11 unknown females, 4 unknown males, 35 white females, 20 white males. 

**Q7. Using this approach determine (i) the average age of wP individuals, (ii) the average age of aP individuals; and (iii) are they significantly different?**

```{r}
library(lubridate)
today()
time_length(today() - ymd("2000-01-01"), "years")
```