---
title: "class10"
format: pdf
prefer-html: true
editor: 
  markdown: 
    wrap: sentence
---

```{r}
url <- "https://raw.githubusercontent.com/fivethirtyeight/data/master/candy-power-ranking/candy-data.csv"
candy_file <- read.csv(url)

candy = data.frame(candy_file, row.names=1)
head(candy)
```

## Q1. How many different candy types are in this dataset?

```{r}
ncol(candy)
```

12. 

    ## Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```

38. 

## Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

Almond Joy!
50.34755.
See below.

## Q4. What is the winpercent value for “Kit Kat”? 

76.7686

## Q5. What is the winpercent value for “Tootsie Roll Snack Bars”? 

49.6535

```{r}
# for twix; example
candy["Twix", ]$winpercent

candy["Almond Joy", ]$winpercent
candy["Kit Kat", ]$winpercent
candy["Tootsie Roll Snack Bars", ]$winpercent
```

## Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

winpercent is insanely high compared to all of the other variables.

## Q7. What do you think a zero and one represent for the candy\$chocolate column? 

0-not chocolate, 1-is chocolate.

```{r}
#install.packages("skimr")
library(skimr)
skim(candy)
```

## Q8. Plot a histogram of winpercent values

See below.

## Q9. Is the distribution of winpercent values symmetrical? 

Not really?
There are multiple modes asymmetrically peaking on the graph, and the data is not evenly scattered on either side.

## Q10. Is the center of the distribution above or below 50%?

Below.
Data ranges from 0-90 on display, and majority of data is below 50.

## Q11. On average is chocolate candy higher or lower ranked than fruit candy? 

Chocolate is higher ranked.

## Q12. Is this difference statistically significant? 

Yes.
p-value = 2.871e-08 \<\<\< 0.05

```{r}
plot(x=candy$winpercent, type="h")

fruitLogical <- candy$winpercent[as.logical(candy$fruity)]
chocLogical <- candy$winpercent[as.logical(candy$chocolate)]
mean(fruitLogical)
mean(chocLogical)

t.test(fruitLogical, chocLogical)
```

## Q13. What are the five least liked candy types in this set?

Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble, Jawbusters.
See below.

## Q14. What are the top 5 all time favorite candy types out of this set? 

Reese's Peanut Butter Cups, Reese's Miniatures, Twix, Kit Kat, Snickers.

```{r}
head(candy[order(candy$winpercent, decreasing=FALSE),], n=5)
head(candy[order(candy$winpercent, decreasing=TRUE),], n=5)
```

## Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
library(ggplot2)
ggplot(candy) + 
  aes(winpercent, rownames(candy)) +
  geom_col()
```

## Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(width=0.7) 
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
```

## Q17. What is the worst ranked chocolate candy?

Charleston Chew

## Q18. What is the best ranked fruity candy? 

Starburst

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols)
```

```{r}
#install.packages("ggrepel")
library(ggrepel)
```

## Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Reese's miniatures

## Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular? 

Nik L Nip

```{r}
# How about a plot of price vs win
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)

ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```

## Q21. Make a barplot again with geom_col() this time using pricepercent and then improve this step by step, first ordering the x-axis by value and finally making a so called “dot chat” or “lollipop” chart by swapping geom_col() for geom_point() + geom_segment().

```{r}
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_segment(aes(yend = reorder(rownames(candy), pricepercent), 
                   xend = 0), col="gray40") +
    geom_point()
```

## Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Chocolate and fruity.
See below.

## Q23. Similarly, what two variables are most positively correlated? 

Chocolate and bar (and winpercent with chocolate)

```{r}
#install.packages("corrplot")
library(corrplot)
cij <- cor(candy)
corrplot(cij)
```

```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
plot(pca$x[,1:2])
```

```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```

```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)
p
```

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

```{r}
#install.packages("plotly")
#library(plotly)
#ggplotly(p)
```

## Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

Fruity, hard, and pluribus.
These do make sense, since fruity candies do tend to be hard and come in packages of multiple candies, which are a better price per individual candy (compared to a single chocolate bar)

```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```
